Parsing Postfix log files

### 2008-02-07 doc/logparser.tex
Added new commands giving the number of log files and number of rules, to make changing easier, and replaced all occurrences of then numbers. The table summarising the most effective restrictions is now automatically generated.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully reconstructed
by the parsing algorithm, including dealing with false negatives and a
discussion of the difficulties in identifying false positives.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote \IP{}
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        93 log files used in testing the parser are shown in
        table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \begin{tabular}[]{lr}
                Restriction type    & Total rejections  \\
                Total rejections    & 5283739           \\
                Unknown recipient   & 2338000           \\
                Unknown sender      & 864148            \\
                \RBL{}s             & 807185            \\
                Greylisting         & 497244            \\
                Non-\FQDN{} HELO    & 403652            \\
            \end{tabular}
        \end{table}

@@

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
complication which occurs only four times in 93 log files tested.}
Requiring changes to the parsing algorithm also complicates upgrades, as
the changes must be preserved during the upgrade, and may clash with
changes made by the developer.  \parsername{} allows the user to add new
rules to the database without changing the parsing algorithm, unless the
new log lines to be parsed require functionality not already provided by
the algorithm.  If the new log lines do require new functionality, new
actions can be added to the parser without modifying existing actions or
other parts of the algorithm; only in the rare case that the new actions
require support from other sections of the code will more extensive changes
be required.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
appendix~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
million lines) were each parsed 10 times, the first run discarded, and the
remaining 9 runs averaged.  The first run is discarded for two reasons:

@@

Rule ordering was mentioned in \sectionref{rule attributes} and will be
covered in greater detail in this section.  At the time of writing there
are 139 different rules, with the top 10\% matching the vast majority of
the log lines, and the remaining log lines split across the other 90\% of
the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming that
the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
after cleanup}.  In the 93 log files used for testing this occurs only five
times in 60,721,709 log lines, but for completeness of the algorithm it
should be dealt with.  The five occurrences in the test log files have the
same characteristics: the \daemon{local} log line showing delivery to a
local mailbox occurs after the \daemon{qmgr} log line showing removal of
the mail from the queue because delivery is completed.  This causes
problems: the mail is not complete, so entry into the database fails; a new
mail is created when the \daemon{local} line is parsed and remains in the
state tables; four warnings are issued per pair of out of order log lines.

@@

The aborted delivery attempts described in
\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the 93
log files used for testing.  The symptoms are the same as in
\sectionref{aborted-delivery-attempts}, except that there \textit{is\/} a
\daemon{cleanup} log line; there does not appear to be anything in the log
file to explain why there are no further log messages.  The only way to
detect these mails is to periodically scan all mails in the state tables,
deleting any mails displaying the following characteristics:

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has 139 rules, fully parsing 93 contiguous log files
from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
coverage is hard because other servers have different restrictions with
custom messages, \RBL{} messages change over time, major releases of
Postfix change warning messages (usually adding more information), etc.,\
so over time the log lines drift and change.  Graph~\refwithpage{rule hits
graph} shows the number of hits for each rule over all 93 log files; it's
obvious that a small number of rules match the vast majority of the lines,
and more than half the rules match fewer than 100 times.

@@

Warnings are issued for any lines which are not parsed; no warnings are
issued for unparsed lines while testing with the 93 test log files, so it
can be safely concluded that there are zero false negatives.  False
positives are harder to quantify: short of examining each of the 60,721,709
log lines and determining which \regex{} parsed it, there is no way to be
sure that every line was parsed by the correct \regex{}, making it
impossible to quantify the false positive rate.  The author is confident
that the false positive rate is extremely low, approaching if not reaching
zero, because the \regexes{} used are as specific as possible and care has
been taken to ensure that there are no unintentionally overlapping
\regexes{}.

@@

At the time of writing the parser is being tested with 93 log files.  There
are 5 warnings produced, but because the parser errs on the side of
producing more warnings rather than fewer those 5 warnings represent 3
instances of 1 problem: 3 connections started before the first log file, so
their initial log entries are missing, leading to warnings when their log
lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the 93 test log files the state tables contain 18
entries, breaking down into:

>>>>

\newcommand{\numberOFlogFILES}[0]{%
    93%
}

\newcommand{\numberOFrules}[0]{%
    142%
}

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 1000 users.  Coverage is described both in terms of
the fraction of lines parsed and the fraction of mails and connections
successfully reconstructed by the parsing algorithm, including dealing with
false negatives and a discussion of the difficulties in identifying false
positives.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote \IP{}
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        \numberOFlogFILES{} log files used in testing the parser are shown
        in table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \input{build/restriction-table-include.tex}
        \end{table}

@@

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
complication which occurs only four times in \numberOFlogFILES{} log files
tested.} Requiring changes to the parsing algorithm also complicates
upgrades, as the changes must be preserved during the upgrade, and may
clash with changes made by the developer.  \parsername{} allows the user to
add new rules to the database without changing the parsing algorithm,
unless the new log lines to be parsed require functionality not already
provided by the algorithm.  If the new log lines do require new
functionality, new actions can be added to the parser without modifying
existing actions or other parts of the algorithm; only in the rare case
that the new actions require support from other sections of the code will
more extensive changes be required.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
deal with 75 MB log files containing 300,000 log lines (for a small mail
server --- large scale mail servers would have much larger log files on a
daily basis).  When generating the data for the graphs included in
appendix~\refwithpage{graphs}, \numberOFlogFILES{} log files (totaling
10.08 GB, 60.72 million lines) were each parsed 10 times, the first run
discarded, and the remaining 9 runs averaged.  The first run is discarded
for two reasons:

@@

Rule ordering was mentioned in \sectionref{rule attributes} and will be
covered in greater detail in this section.  At the time of writing there
are \numberOFrules{} different rules, with the top 10\% matching the vast
majority of the log lines, and the remaining log lines split across the
other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
Assuming that the distribution of log lines is reasonably steady over time,
program efficiency should benefit from trying more frequently matching
rules before those which match less frequently.  To test this hypothesis
three full test runs were performed with different rule orderings:

@@

Occasionally a log file will have out of order log lines which cannot be
dealt with by the techniques described in \sectionref{tracking re-injected
mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
occurs only five times in 60,721,709 log lines, but for completeness of the
algorithm it should be dealt with.  The five occurrences in the test log
files have the same characteristics: the \daemon{local} log line showing
delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
removal of the mail from the queue because delivery is completed.  This
causes problems: the mail is not complete, so entry into the database
fails; a new mail is created when the \daemon{local} line is parsed and
remains in the state tables; four warnings are issued per pair of out of
order log lines.

@@

The aborted delivery attempts described in
\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
delivery attempts described in this section only occur four times in the
\numberOFlogFILES{} log files used for testing.  The symptoms are the same
as in \sectionref{aborted-delivery-attempts}, except that there
\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
anything in the log file to explain why there are no further log messages.
The only way to detect these mails is to periodically scan all mails in the
state tables, deleting any mails displaying the following characteristics:

@@

Full coverage of log lines is easy to achieve yet hard to maintain.  It is
easy to achieve full coverage for a limited set of log files (at the time
of writing the parser has \numberOFrules{} rules, fully parsing
\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
rules are easy to add.  Maintaining full coverage is hard because other
servers have different restrictions with custom messages, \RBL{} messages
change over time, major releases of Postfix change warning messages
(usually adding more information), etc.,\ so over time the log lines drift
and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
for each rule over all \numberOFlogFILES{} log files; it's obvious that a
small number of rules match the vast majority of the lines, and more than
half the rules match fewer than 100 times.

@@

Warnings are issued for any lines which are not parsed; no warnings are
issued for unparsed lines while testing with the \numberOFlogFILES{} test
log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: short of examining each
of the 60,721,709 log lines and determining which \regex{} parsed it, there
is no way to be sure that every line was parsed by the correct \regex{},
making it impossible to quantify the false positive rate.  The author is
confident that the false positive rate is extremely low, approaching if not
reaching zero, because the \regexes{} used are as specific as possible and
care has been taken to ensure that there are no unintentionally overlapping
\regexes{}.

@@

At the time of writing the parser is being tested with \numberOFlogFILES{}
log files.  There are 5 warnings produced, but because the parser errs on
the side of producing more warnings rather than fewer those 5 warnings
represent 3 instances of 1 problem: 3 connections started before the first
log file, so their initial log entries are missing, leading to warnings
when their log lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the \numberOFlogFILES{} test log files the state
tables contain 18 entries, breaking down into:
